---
title: "Australian Real Goods Nowcast — Briefing"
subtitle: "resourcetracker · weekly run"
date: today
format:
  html:
    toc: true
    self-contained: true
    theme: cosmo
  pdf:
    toc: true
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
params:
  outputs_dir: "../../outputs"
  warehouse:  "../../data/warehouse.duckdb"
---

```{r setup}
library(dplyr)
library(readr)
library(ggplot2)
library(fs)
library(DBI)
library(duckdb)

outputs  <- params$outputs_dir
wh_path  <- params$warehouse

nowcast    <- read_csv(path(outputs, "nowcast_current.csv"),
                       show_col_types = FALSE)
tonnage    <- read_csv(path(outputs, "tonnage_monthly.csv"),
                       show_col_types = FALSE)
diag_fit   <- read_csv(path(outputs, "bridge_diagnostics.csv"),
                       show_col_types = FALSE)

con <- tryCatch(
  dbConnect(duckdb(), dbdir = wh_path, read_only = TRUE),
  error = function(e) NULL
)

pull_safely <- function(sql) {
  if (is.null(con)) return(tibble())
  tryCatch(as_tibble(dbGetQuery(con, sql)), error = function(e) tibble())
}

history   <- pull_safely("SELECT * FROM mart.nowcast_history ORDER BY run_timestamp")
abs_5302  <- pull_safely("SELECT * FROM raw.abs_5302_quarterly ORDER BY quarter_end")
```

## Headline nowcast

```{r headline}
if (nrow(nowcast) == 0 || is.na(nowcast$point_estimate[1])) {
  cat("*No nowcast available this run.*")
} else {
  nc <- nowcast[1, ]
  cat(sprintf(
    "**Quarter %s:** A$%s million (80%% CI: A$%s – A$%s; 95%% CI: A$%s – A$%s). Share of quarter observed: %.0f%%.",
    format(nc$quarter_end, "%b %Y"),
    formatC(nc$point_estimate, format = "d", big.mark = ","),
    formatC(nc$lower_80,       format = "d", big.mark = ","),
    formatC(nc$upper_80,       format = "d", big.mark = ","),
    formatC(nc$lower_95,       format = "d", big.mark = ","),
    formatC(nc$upper_95,       format = "d", big.mark = ","),
    100 * nc$share_observed
  ))
}
```

```{r headline-plot, fig.width=7, fig.height=3.5}
if (nrow(history) > 0) {
  history |>
    mutate(quarter_end = as.Date(quarter_end)) |>
    ggplot(aes(run_timestamp, point_estimate)) +
    geom_ribbon(aes(ymin = lower_95, ymax = upper_95), alpha = 0.15) +
    geom_ribbon(aes(ymin = lower_80, ymax = upper_80), alpha = 0.25) +
    geom_line() +
    labs(x = "Run timestamp",
         y = "Real goods credits (A$m, chain-volume)",
         title = "Nowcast evolution within the current quarter") +
    theme_minimal()
}
```

## Commodity decomposition

```{r decomposition}
if (nrow(tonnage) > 0) {
  tonnage |>
    filter(!is.na(commodity)) |>
    group_by(commodity) |>
    summarise(tonnage = sum(tonnage, na.rm = TRUE), .groups = "drop") |>
    mutate(share = tonnage / sum(tonnage)) |>
    arrange(desc(share)) |>
    knitr::kable(digits = c(NA, 0, 3),
                 col.names = c("Commodity", "Tonnage (summed)", "Share"))
} else {
  cat("*Tonnage panel empty.*")
}
```

## Change since last run

```{r delta}
if (nrow(history) >= 2) {
  current <- history[nrow(history), ]
  prev    <- history[nrow(history) - 1, ]
  tibble(
    metric = c("Point estimate", "80% lower", "80% upper",
               "95% lower", "95% upper", "Share observed"),
    prior  = c(prev$point_estimate, prev$lower_80, prev$upper_80,
               prev$lower_95, prev$upper_95, prev$share_observed),
    now    = c(current$point_estimate, current$lower_80, current$upper_80,
               current$lower_95, current$upper_95, current$share_observed),
    delta  = c(current$point_estimate - prev$point_estimate,
               current$lower_80       - prev$lower_80,
               current$upper_80       - prev$upper_80,
               current$lower_95       - prev$lower_95,
               current$upper_95       - prev$upper_95,
               current$share_observed - prev$share_observed)
  ) |>
    knitr::kable(digits = c(NA, 1, 1, 1))
} else {
  cat("*First run this quarter — no prior run to compare against.*")
}
```

## Comparison to ABS latest actual

```{r abs-compare}
if (nrow(abs_5302) > 0) {
  latest_abs <- abs_5302 |>
    filter(!is.na(value_chainvol_aud_m), value_chainvol_aud_m > 0) |>
    arrange(desc(quarter_end)) |>
    slice(1)
  cat(sprintf(
    "Latest ABS 5302.0 actual (quarter ending %s): **A$%s million**.\n\n",
    format(as.Date(latest_abs$quarter_end), "%b %Y"),
    formatC(latest_abs$value_chainvol_aud_m,
            format = "d", big.mark = ",")
  ))
} else {
  cat("*ABS 5302.0 panel empty.*")
}
```

## Flagged anomalies

```{r anomalies-table}
anoms <- pull_safely(
  "SELECT obs_date, port_id, commodity, tonnage, z_score
   FROM mart.latest_anomalies
   ORDER BY ABS(z_score) DESC LIMIT 10"
)
if (nrow(anoms) > 0) {
  knitr::kable(anoms, digits = c(NA, NA, NA, 0, 2))
} else {
  cat("*No >2σ departures flagged in the last 28 days.*")
}
```

## Methodology footnote

Bridge regression per commodity on log levels with an AR(1) lag:

$$\log y_{c,m} = \beta_0 + \beta_1 \log T^{SA}_{c,m} + \beta_2 \log P_{c,m} + \beta_3 \log y_{c,m-1} + \varepsilon_{c,m}$$

Fit on 2019-01 → 2023-12; validated on 2024+. Newey-West SEs, lag 3.
Quarterly aggregate converted to chain-volume via the ABS 5302.0
implicit deflator (trailing 4-quarter mean as forecast). Uncertainty
bands via 1,000 bootstrap draws of bridge residuals, scaled by
$\sqrt{1 - \text{share\_observed}}$.

```{r close-con}
if (!is.null(con)) dbDisconnect(con, shutdown = TRUE)
```
