---
title: "Australian Resource Export Volumes"
subtitle: "Quarterly nowcast — iron ore and coal"
date: "`r format(Sys.Date(), '%e %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    self_contained: true
    theme: null
    highlight: null
    css: style.css
    mathjax: default
    md_extensions: "+raw_html"
params:
  outputs_dir: "../../outputs"
  warehouse_dir: "../../data/warehouse"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                     dev = "png", dpi = 144)
library(dplyr)
library(readr)
library(ggplot2)
library(fs)

# Unified palette & ggplot theme so charts feel like part of the page,
# not a foreign bitmap.
.palette <- list(
  text       = "#1A1F2C",
  text_muted = "#5A6478",
  border     = "#E4E1D9",
  accent     = "#1C3D5A",
  accent2    = "#8A4A1F",
  grid       = "#EFEDE7",
  bg         = "#FAFAF7"
)
theme_rt <- function(base_size = 11) {
  theme_minimal(base_family = "Inter", base_size = base_size) +
    theme(
      plot.background  = element_rect(fill = .palette$bg, colour = NA),
      panel.background = element_rect(fill = .palette$bg, colour = NA),
      panel.grid.major = element_line(colour = .palette$grid, size = 0.4),
      panel.grid.minor = element_blank(),
      axis.text  = element_text(colour = .palette$text_muted),
      axis.title = element_text(colour = .palette$text_muted, size = base_size - 1),
      axis.line.x = element_line(colour = .palette$border, size = 0.5),
      axis.ticks  = element_line(colour = .palette$border, size = 0.4),
      strip.text  = element_text(colour = .palette$text, face = "bold",
                                 hjust = 0, size = base_size + 1),
      strip.background = element_blank(),
      plot.title = element_text(colour = .palette$text, face = "bold",
                                size = base_size + 2,
                                margin = margin(0, 0, 8, 0)),
      plot.subtitle = element_text(colour = .palette$text_muted,
                                   size = base_size - 1,
                                   margin = margin(0, 0, 12, 0)),
      plot.caption  = element_text(colour = .palette$text_muted,
                                   size = base_size - 2, hjust = 0,
                                   margin = margin(12, 0, 0, 0)),
      legend.position = "none"
    )
}
theme_set(theme_rt())
.commodity_colours <- c(iron_ore     = .palette$accent,
                         coal         = .palette$accent2,
                         coal_met     = .palette$accent2,
                         coal_thermal = "#3F8A6A")    # muted teal-green
.commodity_labels  <- c(iron_ore     = "Iron ore",
                         coal         = "Coal",
                         coal_met     = "Metallurgical coal",
                         coal_thermal = "Thermal coal")

outputs <- params$outputs_dir
wh_dir  <- params$warehouse_dir

safe_read_csv <- function(p) {
  if (!file_exists(p)) return(tibble())
  suppressMessages(read_csv(p, show_col_types = FALSE))
}
safe_read_wh <- function(name) {
  p <- path(wh_dir, paste0(name, ".rds"))
  if (!file_exists(p)) return(tibble())
  tibble::as_tibble(readRDS(p))
}

nowcast    <- safe_read_csv(path(outputs, "nowcast_current.csv"))
chain_vol  <- safe_read_csv(path(outputs, "nowcast_chain_vol.csv"))
tonnage    <- safe_read_csv(path(outputs, "tonnage_quarterly.csv"))
diag_fit   <- safe_read_csv(path(outputs, "bridge_diagnostics.csv"))
history   <- safe_read_wh("mart_nowcast_history")
disr      <- safe_read_wh("raw_disr_req_quarterly")
anomalies <- safe_read_wh("mart_latest_anomalies")
ports_dim <- safe_read_wh("mart_dim_port")

com_levels <- c("iron_ore", "coal_met", "coal_thermal", "coal")

# For each commodity in `nowcast`, find the most-recent prior row in
# `history` for the same quarter that isn't this run, and compute the
# point + 80% band delta. Returns a tibble with one row per commodity;
# delta cols are NA when no prior exists.
compute_wow_delta <- function(nowcast, history) {
  if (nrow(nowcast) == 0 || nrow(history) == 0) {
    return(tibble::tibble(commodity = character(),
                          delta_point = double(),
                          delta_pct = double(),
                          prior_run = as.POSIXct(NA),
                          hours_since = double()))
  }
  history <- history |>
    mutate(run_timestamp = as.POSIXct(run_timestamp),
           quarter_end   = as.Date(quarter_end))
  purrr::map_dfr(seq_len(nrow(nowcast)), function(i) {
    cur <- nowcast[i, ]
    cur_ts <- as.POSIXct(cur$run_timestamp)
    prior <- history |>
      filter(commodity == cur$commodity,
             quarter_end == as.Date(cur$quarter_end),
             run_timestamp < cur_ts) |>
      arrange(desc(run_timestamp)) |>
      slice(1)
    if (nrow(prior) == 0) {
      tibble::tibble(commodity   = cur$commodity,
                     delta_point = NA_real_,
                     delta_pct   = NA_real_,
                     prior_run   = as.POSIXct(NA),
                     hours_since = NA_real_)
    } else {
      d <- cur$point_estimate_Mt - prior$point_estimate
      tibble::tibble(
        commodity   = cur$commodity,
        delta_point = d,
        delta_pct   = 100 * d / prior$point_estimate,
        prior_run   = prior$run_timestamp,
        hours_since = as.numeric(difftime(cur_ts, prior$run_timestamp,
                                          units = "hours"))
      )
    }
  })
}

# WoW delta only meaningful for the in-quarter nowcast (h = 0); the
# Q+1 forecast has no rolling-comparison anchor that's stable enough
# to be useful.
nowcast_h0 <- if ("horizon" %in% names(nowcast)) {
  dplyr::filter(nowcast, horizon == 0L)
} else nowcast
history_h0 <- if (!is.null(history) && nrow(history) > 0 && "horizon" %in% names(history)) {
  dplyr::filter(history, horizon == 0L)
} else history
delta <- compute_wow_delta(nowcast_h0, history_h0)
```

## National accounts chain-volume — latest unpublished quarter

```{r chain-vol-backcast, results='asis'}
cv_label <- function(x) {
  switch(x,
         iron_ore   = "Iron ore",
         coal_total = "Coal (total)",
         tools::toTitleCase(gsub("_", " ", x)))
}

fmt_am <- function(x) format(round(x), big.mark = ",")

if (nrow(chain_vol) == 0) {
  cat('<div class="callout"><span class="label">Status</span>',
      'No chain-volume nowcast — ABS Balance of Payments data is unavailable.',
      '</div>\n')
} else {
  bc <- chain_vol |>
    filter(horizon < 0L) |>
    mutate(commodity = factor(commodity,
                              levels = c("iron_ore", "coal_total"))) |>
    arrange(commodity, quarter_end)

  if (nrow(bc) == 0) {
    cat('<div class="callout"><span class="label">Status</span>',
        'No gap quarters — ABS chain-volume is current through the latest',
        'PortWatch-observed quarter.</div>\n')
  } else {
    qtr_label <- format(as.Date(bc$quarter_end[1]), "%B %Y")
    cat(sprintf(
      '<p class="kicker">Backcast for %s &mdash; ABS BoP not yet published</p>\n',
      qtr_label
    ))
    cat('<p style="color:var(--text-muted);max-width:680px;margin:0 0 1rem">',
        'ABS Balance of Payments chain-volume measures lag actual quarter-end',
        'by ~2 months. These estimates anticipate the upcoming release by',
        'applying the validated growth-rate pass-through bridge: YoY growth in',
        'DISR physical tonnes (back-cast from complete PortWatch data) is',
        'chained onto the most recent ABS chain-volume level.',
        '</p>\n')
    cat('<div class="headline-grid">\n')
    for (i in seq_len(nrow(bc))) {
      r <- bc[i, ]
      growth_pct <- 100 * (r$growth_factor - 1)
      growth_class <- if (growth_pct >  0.5) "up"
                      else if (growth_pct < -0.5) "down"
                      else "flat"
      growth_sign <- if (growth_pct >= 0) "+" else "&minus;"
      card_html <- sprintf(
        '<div class="headline-card"><span class="label">%s</span><div class="value-row"><span class="value">%s</span><span class="unit">A&#36;m</span></div><span class="ci">80%% CI &middot; %s &ndash; %s A&#36;m</span><span class="delta %s">%s%.1f%% YoY</span></div>',
        cv_label(as.character(r$commodity)),
        fmt_am(r$point_estimate_Am),
        fmt_am(r$lower_80_Am), fmt_am(r$upper_80_Am),
        growth_class, growth_sign, abs(growth_pct)
      )
      cat(card_html, "\n", sep = "")
    }
    cat('</div>\n')
  }

  # Q2 nowcast + Q3 forecast — secondary cards below the backcast.
  fwd <- chain_vol |>
    filter(horizon >= 0L) |>
    mutate(commodity = factor(commodity,
                              levels = c("iron_ore", "coal_total"))) |>
    arrange(horizon, commodity)

  if (nrow(fwd) > 0) {
    cat('<h3 style="margin-top:2rem">Looking ahead &mdash; chain-volume</h3>\n')
    cat('<p style="color:var(--text-muted);max-width:680px;margin:0 0 1rem">',
        'Current-quarter nowcast (partial PortWatch coverage) and next-quarter',
        'forecast (no PortWatch yet). Bands widen with horizon as the tonnage',
        'extrapolation contributes more uncertainty.',
        '</p>\n')
    cat('<div class="headline-grid">\n')
    for (i in seq_len(nrow(fwd))) {
      r <- fwd[i, ]
      type_lbl <- if (r$horizon == 0L) "nowcast" else "forecast"
      qlbl <- format(as.Date(r$quarter_end), "%b %Y")
      growth_pct <- 100 * (r$growth_factor - 1)
      growth_sign <- if (growth_pct >= 0) "+" else "&minus;"
      card_html <- sprintf(
        '<div class="headline-card"><span class="label">%s &middot; %s %s</span><div class="value-row"><span class="value">%s</span><span class="unit">A&#36;m</span></div><span class="ci">80%% CI &middot; %s &ndash; %s</span><span class="delta flat">%s%.1f%% YoY</span></div>',
        cv_label(as.character(r$commodity)),
        qlbl, type_lbl,
        fmt_am(r$point_estimate_Am),
        fmt_am(r$lower_80_Am), fmt_am(r$upper_80_Am),
        growth_sign, abs(growth_pct)
      )
      cat(card_html, "\n", sep = "")
    }
    cat('</div>\n')
  }
}
```

## Physical tonnage nowcasts

```{r top-line, results='asis'}
commodity_label <- function(x) {
  switch(x,
         iron_ore     = "Iron ore",
         coal         = "Coal",
         coal_met     = "Metallurgical coal",
         coal_thermal = "Thermal coal",
         tools::toTitleCase(gsub("_", " ", x)))
}

if (nrow(nowcast) == 0) {
  cat('<div class="callout"><span class="label">Status</span>',
      'No nowcast produced this run.</div>')
} else {
  # Headline cards show the current quarter (h = 0). The next-quarter
  # outlook (h = 1) renders in its own section below.
  if (!"horizon" %in% names(nowcast)) nowcast$horizon <- 0L
  nc <- nowcast |>
    filter(horizon == 0L) |>
    mutate(commodity = factor(commodity, levels = com_levels)) |>
    arrange(commodity) |>
    left_join(delta, by = "commodity")
  share_pct <- round(100 * nc$share_observed[1])
  qtr_label <- format(as.Date(nc$quarter_end[1]), "%B %Y")

  cat(sprintf(
    '<p class="kicker">Current quarter — %s · %d%% of the quarter observed</p>\n',
    qtr_label, share_pct
  ))
  cat('<div class="headline-grid">\n')
  for (i in seq_len(nrow(nc))) {
    r <- nc[i, ]
    delta_class <- if (is.na(r$delta_point)) "flat"
                   else if (r$delta_point >  0.05) "up"
                   else if (r$delta_point < -0.05) "down"
                   else "flat"
    delta_text <- if (is.na(r$delta_point)) {
      "First nowcast for this quarter"
    } else {
      sign_lbl  <- if (r$delta_point >= 0) "+" else "−"
      sign_pct  <- if (r$delta_pct   >= 0) "+" else "−"
      hours_lbl <- if (r$hours_since < 24) sprintf("%.0fh ago", r$hours_since)
                   else sprintf("%.0f days ago", r$hours_since / 24)
      sprintf("%s%.1f Mt&nbsp;&nbsp;(%s%.1f%%) vs the run %s",
              sign_lbl, abs(r$delta_point),
              sign_pct, abs(r$delta_pct),
              hours_lbl)
    }
    # Single-line HTML so pandoc treats the whole card as one raw-HTML
    # block and doesn't paragraph-wrap the inline spans.
    card_html <- sprintf(
      '<div class="headline-card"><span class="label">%s</span><div class="value-row"><span class="value">%.1f</span><span class="unit">Mt</span></div><span class="ci">80%% CI · %.1f – %.1f Mt</span><span class="delta %s">%s</span></div>',
      commodity_label(as.character(r$commodity)),
      r$point_estimate_Mt,
      r$lower_80, r$upper_80,
      delta_class, delta_text
    )
    cat(card_html, "\n", sep = "")
  }
  cat('</div>\n')
}
```

```{r attribution, results='asis'}
# Mechanical decomposition of each in-quarter nowcast into additive
# contributions: lag-4 baseline + tonnage signal + lag4 drift +
# intercept + other regressors (e.g. price for price_aug). Surfaces
# *why* the nowcast differs from a naive "same quarter last year"
# benchmark. Source: `decompose_nowcast()` in R/attribution.R.
attr_cols <- c("v_lag4_Mt", "tonnage_signal_Mt", "lag4_drift_Mt",
               "intercept_Mt", "other_Mt")
if (nrow(nowcast) > 0 && all(attr_cols %in% names(nowcast))) {
  attr_h0 <- nowcast |>
    dplyr::filter(horizon == 0L, !is.na(v_lag4_Mt))
  if (nrow(attr_h0) > 0) {
    fmt_signed_mt <- function(x) {
      if (!is.finite(x)) return("—")
      sprintf(if (x >= 0) "+%.1f" else "%.1f", x)
    }
    rows <- attr_h0 |>
      dplyr::mutate(
        commodity     = factor(commodity, levels = com_levels),
        anchor_adj_Mt = intercept_Mt + lag4_drift_Mt
      ) |>
      dplyr::arrange(commodity) |>
      dplyr::transmute(
        commodity   = vapply(as.character(commodity), commodity_label,
                             character(1)),
        `last year (V_{Q-4})` = sprintf("%.1f Mt", v_lag4_Mt),
        `tonnage signal`      = vapply(tonnage_signal_Mt, fmt_signed_mt,
                                       character(1)),
        `model anchor adj.`   = vapply(anchor_adj_Mt,     fmt_signed_mt,
                                       character(1)),
        `other`               = vapply(other_Mt,          fmt_signed_mt,
                                       character(1)),
        `= nowcast`           = sprintf("%.1f Mt", point_estimate_Mt)
      )
    cat('<h3 style="margin-top:1.5rem">Why the nowcast says what it says</h3>\n')
    cat('<p style="color:var(--text-muted);max-width:680px;margin:0 0 0.75rem">',
        'Each headline above broken into additive contributions vs the',
        'seasonal-naive baseline (same quarter last year). All values',
        'in Mt; rows sum to the nowcast up to a small non-linearity in',
        '<code>exp(Δlog) − 1</code>.',
        '"Model anchor adj." combines the intercept and the lag-4',
        'coefficient drift &mdash; effectively zero for <code>bojo</code>',
        'specs (which impose β<sub>4</sub> = 1), small for the others.',
        '"Other" fires only when a price-augmented spec wins production',
        'for a commodity.',
        '</p>\n')
    print(knitr::kable(rows, format = "html", escape = FALSE,
                       table.attr = 'class="attribution-table"'))
  }
}
```

```{r next-quarter, results='asis'}
# Next-quarter forecast (h = 1) -- pure seasonal forecast for unobserved
# tonnage, fed through the same bridge. Much wider error bars by
# construction; presented separately so the headline cards stay focused
# on the in-quarter nowcast.
if ("horizon" %in% names(nowcast) && any(nowcast$horizon == 1L)) {
  nq <- nowcast |>
    filter(horizon == 1L) |>
    mutate(commodity = factor(commodity, levels = com_levels)) |>
    arrange(commodity)
  qlbl <- format(as.Date(nq$quarter_end[1]), "%B %Y")
  cat(sprintf('<h3 style="margin-top:2rem">Next-quarter outlook &middot; %s</h3>\n', qlbl))
  cat('<p style="color:var(--text-muted);max-width:640px;margin:0 0 1rem">',
      'Forecast for the quarter after the headline above, using seasonal',
      'PortWatch patterns + the most recent observed pace. Bands are',
      'wider by construction (no in-quarter observations to shrink',
      'them).</p>\n')
  cat('<div class="headline-grid">\n')
  for (i in seq_len(nrow(nq))) {
    r <- nq[i, ]
    card_html <- sprintf(
      '<div class="headline-card"><span class="label">%s &middot; forecast</span><div class="value-row"><span class="value">%.1f</span><span class="unit">Mt</span></div><span class="ci">80%% CI · %.1f – %.1f Mt</span></div>',
      commodity_label(as.character(r$commodity)),
      r$point_estimate_Mt,
      r$lower_80, r$upper_80
    )
    cat(card_html, "\n", sep = "")
  }
  cat('</div>\n')
}
```

## Nowcast vs recent actuals

```{r actuals-plot, fig.width=9, fig.height=4.5}
if (nrow(disr) > 0 && nrow(nowcast) > 0) {
  hist <- disr |>
    mutate(quarter_end = as.Date(quarter_end),
           commodity   = factor(commodity, levels = com_levels)) |>
    filter(!is.na(commodity),
           quarter_end >= as.Date("2019-01-01")) |>
    arrange(commodity, quarter_end)

  if (!"horizon" %in% names(nowcast)) nowcast$horizon <- 0L
  nc <- nowcast |>
    filter(horizon == 0L) |>
    mutate(quarter_end = as.Date(quarter_end),
           commodity   = factor(commodity, levels = com_levels))
  nc_fwd <- nowcast |>
    filter(horizon == 1L) |>
    mutate(quarter_end = as.Date(quarter_end),
           commodity   = factor(commodity, levels = com_levels))

  p <- ggplot(hist, aes(quarter_end, tonnes_Mt)) +
    geom_line(colour = .palette$text_muted, linewidth = 0.5) +
    geom_point(colour = .palette$text_muted, size = 1.2) +
    geom_linerange(data = nc,
                   aes(x = quarter_end,
                       ymin = lower_95, ymax = upper_95,
                       y = point_estimate_Mt),
                   colour = .palette$accent, alpha = 0.25, linewidth = 1.4) +
    geom_linerange(data = nc,
                   aes(x = quarter_end,
                       ymin = lower_80, ymax = upper_80,
                       y = point_estimate_Mt),
                   colour = .palette$accent, alpha = 0.55, linewidth = 2.2) +
    geom_point(data = nc,
               aes(quarter_end, point_estimate_Mt),
               colour = .palette$accent, size = 3.4)
  if (nrow(nc_fwd) > 0) {
    p <- p +
      geom_linerange(data = nc_fwd,
                     aes(x = quarter_end,
                         ymin = lower_80, ymax = upper_80,
                         y = point_estimate_Mt),
                     colour = .palette$accent, alpha = 0.35, linewidth = 1.6) +
      geom_point(data = nc_fwd,
                 aes(quarter_end, point_estimate_Mt),
                 shape = 21, fill = .palette$bg,
                 colour = .palette$accent, size = 3.4, stroke = 1.2)
  }
  p +
    facet_wrap(~ commodity, scales = "free_y",
               labeller = labeller(commodity = .commodity_labels)) +
    labs(x = NULL, y = "Quarterly exports (Mt)",
         title = "DISR quarterly export volume, with current and next-quarter nowcasts",
         subtitle = "Grey: published DISR T16 actuals. Filled blue: current-quarter nowcast (80/95% bands). Hollow blue: next-quarter forecast (80% band).")
}
```

## Forecast vs actual — closed quarters

```{r forecast-vs-actual, results='asis'}
# For each commodity × quarter that DISR has now published, find the
# *final pre-publication* nowcast (the latest h=0 run we made before
# DISR's release ~5 weeks after quarter close) and report the error.
# This closes the validation loop the project brief asked for:
#   "When March 2026 ITG is released by ABS in early May, I can
#    compare my April 1 nowcast to the ABS actual."
DISR_LAG_DAYS <- 42L  # DISR REQ T16 releases ~5-6 weeks after quarter close

if (nrow(history) == 0 || nrow(disr) == 0) {
  cat("*No closed quarters with prior nowcasts to compare yet.*\n")
} else {
  hist_h0 <- history |>
    dplyr::mutate(
      run_timestamp = as.POSIXct(run_timestamp),
      quarter_end   = as.Date(quarter_end),
      horizon       = dplyr::coalesce(as.integer(horizon), 0L)
    ) |>
    dplyr::filter(horizon == 0L)

  actuals <- disr |>
    dplyr::mutate(quarter_end = as.Date(quarter_end)) |>
    dplyr::transmute(commodity, quarter_end, actual_Mt = tonnes_Mt)

  # For each (commodity, quarter_end), keep the last nowcast whose run
  # landed within DISR_LAG_DAYS of quarter close — i.e., the call made
  # while the quarter was fresh but before the official number was out.
  joined <- hist_h0 |>
    dplyr::inner_join(actuals, by = c("commodity", "quarter_end")) |>
    dplyr::mutate(
      days_after_qend = as.numeric(difftime(as.Date(run_timestamp),
                                            quarter_end, units = "days"))
    ) |>
    dplyr::filter(days_after_qend <= DISR_LAG_DAYS) |>
    dplyr::arrange(commodity, quarter_end, dplyr::desc(run_timestamp)) |>
    dplyr::group_by(commodity, quarter_end) |>
    dplyr::slice(1) |>
    dplyr::ungroup() |>
    dplyr::filter(quarter_end < Sys.Date())  # only closed quarters

  if (nrow(joined) == 0) {
    cat(sprintf(
      "*No closed quarters with a prior in-window nowcast yet. Comparisons appear here once a quarter we nowcasted within %d days of close has been published by DISR.*\n",
      DISR_LAG_DAYS
    ))
  } else {
    joined |>
      dplyr::mutate(
        commodity   = factor(commodity, levels = com_levels),
        nowcast_Mt  = round(point_estimate, 1),
        actual_Mt_r = round(actual_Mt, 1),
        error_Mt    = round(point_estimate - actual_Mt, 2),
        error_pct   = round(100 * (point_estimate - actual_Mt) / actual_Mt, 1),
        inside_80   = ifelse(actual_Mt >= lower_80 & actual_Mt <= upper_80,
                             "yes", "no"),
        inside_95   = ifelse(actual_Mt >= lower_95 & actual_Mt <= upper_95,
                             "yes", "no")
      ) |>
      dplyr::arrange(commodity, quarter_end) |>
      dplyr::select(commodity, quarter_end,
                    `nowcast (Mt)` = nowcast_Mt,
                    `actual (Mt)`  = actual_Mt_r,
                    `error (Mt)`   = error_Mt,
                    `error (%)`    = error_pct,
                    `in 80% CI`    = inside_80,
                    `in 95% CI`    = inside_95) |>
      knitr::kable(
        caption = paste(
          "Forecast accuracy on quarters that have closed and been",
          "published by DISR. Each row is the last nowcast we made for",
          sprintf("that quarter within %d days of close — i.e., the",
                  DISR_LAG_DAYS),
          "call we'd have stood behind before the official number landed.",
          "`error = nowcast - actual` (Mt and %)."
        )
      )

    # Per-commodity summary: count, mean abs error, hit-rate at 80/95.
    summary_df <- joined |>
      dplyr::group_by(commodity) |>
      dplyr::summarise(
        n          = dplyr::n(),
        mae_Mt     = mean(abs(point_estimate - actual_Mt)),
        mape_pct   = 100 * mean(abs(point_estimate - actual_Mt) / actual_Mt),
        hit_80_pct = 100 * mean(actual_Mt >= lower_80 & actual_Mt <= upper_80),
        hit_95_pct = 100 * mean(actual_Mt >= lower_95 & actual_Mt <= upper_95),
        .groups    = "drop"
      ) |>
      dplyr::mutate(
        commodity  = factor(commodity, levels = com_levels),
        mae_Mt     = round(mae_Mt, 2),
        mape_pct   = round(mape_pct, 1),
        hit_80_pct = round(hit_80_pct, 0),
        hit_95_pct = round(hit_95_pct, 0)
      ) |>
      dplyr::arrange(commodity)

    summary_df |>
      dplyr::select(commodity,
                    `n (quarters)` = n,
                    `MAE (Mt)`     = mae_Mt,
                    `MAPE (%)`     = mape_pct,
                    `80% CI hit-rate (%)` = hit_80_pct,
                    `95% CI hit-rate (%)` = hit_95_pct) |>
      knitr::kable(
        caption = paste(
          "Per-commodity summary across the closed-quarter sample above.",
          "Hit-rates should approach the nominal band coverage (80 / 95%)",
          "if the bootstrap bands are well-calibrated — see the",
          "calibration section below for the formal check."
        )
      )
  }
}
```

## Band calibration

```{r calibration, results='asis'}
# Empirical coverage of the bootstrap 80/95% bands when reconstructed
# at share_observed = 0 for each backtest quarter (computed by
# `backtest_coverage()` and joined into bridge_diagnostics.csv).
# We compare against the nominal 80 and 95% targets. Material gaps
# point at residual-bootstrap miscalibration — the principled fix is
# conformal prediction (see docs/ROADMAP.md backlog #9).
cov_cols <- c("coverage_80", "coverage_95")
if (nrow(diag_fit) > 0 && all(cov_cols %in% names(diag_fit))) {
  cal <- diag_fit |>
    dplyr::filter(!is.na(coverage_80) | !is.na(coverage_95)) |>
    dplyr::transmute(
      commodity   = factor(commodity, levels = com_levels),
      spec,
      n_oos,
      coverage_80,
      coverage_95,
      production  = ifelse(isTRUE(production_choice) |
                           production_choice %in% c("TRUE","true"),
                           "★", "")
    ) |>
    dplyr::arrange(commodity, spec)
  if (nrow(cal) == 0) {
    cat("*No coverage data yet — needs at least one backtest quarter per spec.*\n")
  } else {
    cal |>
      dplyr::mutate(
        `coverage 80% (target 0.80)` = sprintf("%.2f", coverage_80),
        `coverage 95% (target 0.95)` = sprintf("%.2f", coverage_95)
      ) |>
      dplyr::select(`★` = production, commodity, spec,
                    `n (OOS)` = n_oos,
                    `coverage 80% (target 0.80)`,
                    `coverage 95% (target 0.95)`) |>
      knitr::kable(
        caption = paste(
          "Empirical band coverage on backtest quarters, reconstructed",
          "at share_observed = 0 (full residual bootstrap, no in-quarter",
          "shrink). Well-calibrated bands produce values near 0.80 and",
          "0.95. Combination specs (`equal_avg`, `inv_mse`) lack a",
          "single lm fit and are omitted."
        )
      )
  }
}
```

## Bridge diagnostics

```{r diagnostics}
if (nrow(diag_fit) > 0) {
  diag_fit |>
    mutate(
      commodity      = factor(commodity, levels = com_levels),
      production     = ifelse(isTRUE(production_choice) |
                              production_choice %in% c("TRUE","true"),
                              "★", ""),
      beta_tonnage   = ifelse(is.na(beta_tonnage), "",
                              sprintf("%.3f (se %.3f)",
                                      beta_tonnage, beta_tonnage_se)),
      beta_lag4      = ifelse(is.na(beta_lag4), "",
                              sprintf("%.3f", beta_lag4)),
      lag4_eq_1_pval = ifelse(is.na(beta_lag4_eq_1_pval), "",
                              sprintf("%.3f", beta_lag4_eq_1_pval)),
      chow_pval_fmt  = ifelse(is.na(chow_pval), "",
                              sprintf("%.2f", chow_pval)),
      r2_train       = ifelse(is.na(r_squared), "",
                              sprintf("%.2f", r_squared)),
      r2_oos         = ifelse(is.na(r_squared_oos), "",
                              sprintf("%.2f", r_squared_oos)),
      rmse_valid     = ifelse(is.na(rmse_valid), "",
                              sprintf("%.2f", rmse_valid)),
      ratio_vs_naive = ifelse(is.na(ratio_vs_naive), "",
                              sprintf("%.2f", ratio_vs_naive))
    ) |>
    arrange(commodity, spec) |>
    select(`★` = production,
           commodity, spec,
           `R²` = r2_train, `R²_oos` = r2_oos,
           `β_T (se)` = beta_tonnage,
           `β_lag4` = beta_lag4,
           `p(β_lag4=1)` = lag4_eq_1_pval,
           `p(Chow)` = chow_pval_fmt,
           rmse_valid,
           `RMSE/naive` = ratio_vs_naive) |>
    knitr::kable(
      caption = paste(
        "★ marks the production-deployed spec per commodity (best of the",
        "candidate set + combinations by backtest RMSE). `R²` is",
        "in-sample, `R²_oos` is walk-forward out-of-sample.",
        "`p(β_lag4=1)` is the Wald-test p-value for the YoY-on-YoY",
        "restriction (NA for `bojo`, which imposes it; `equal_avg` and",
        "`inv_mse` are combination forecasts and have no in-sample β).",
        "`p(Chow)` tests for a structural break at the midpoint of the",
        "training sample; values < 0.05 suggest the AIS↔DISR relationship",
        "shifted within the fit window.",
        "`RMSE/naive` < 1 means the spec beats the seasonal random walk."
      )
    )
}
```

## Backtest — every spec, every quarter

```{r backtest-viz, fig.width=9, fig.height=4.5}
# Per-commodity walk-forward backtest: actuals plotted against every
# candidate spec's point forecast for each validation quarter. The
# production-deployed spec is emphasised; the others are faded so the
# choice is visible at a glance.
backtest_rds <- safe_read_wh("derived_backtest_results")
prod_picks <- if (nrow(diag_fit) > 0 && "production_choice" %in% names(diag_fit)) {
  diag_fit |>
    dplyr::mutate(production_choice = as.character(production_choice)) |>
    dplyr::filter(production_choice %in% c("TRUE","true")) |>
    dplyr::select(commodity, prod_spec = spec)
} else tibble::tibble(commodity = character(), prod_spec = character())

if (nrow(backtest_rds) > 0) {
  bt <- backtest_rds |>
    dplyr::mutate(
      commodity   = factor(commodity, levels = com_levels),
      quarter_end = as.Date(quarter_end)
    ) |>
    dplyr::filter(!is.na(point_estimate)) |>
    dplyr::left_join(prod_picks, by = "commodity") |>
    dplyr::mutate(is_production = !is.na(prod_spec) & spec == prod_spec)

  actuals_bt <- bt |>
    dplyr::distinct(commodity, quarter_end, actual)

  ggplot() +
    geom_line(data = actuals_bt,
              aes(quarter_end, actual),
              colour = .palette$text, linewidth = 0.7) +
    geom_point(data = actuals_bt,
               aes(quarter_end, actual),
               colour = .palette$text, size = 1.6) +
    geom_line(data = dplyr::filter(bt, !is_production),
              aes(quarter_end, point_estimate, group = spec),
              colour = .palette$text_muted, alpha = 0.35, linewidth = 0.4) +
    geom_line(data = dplyr::filter(bt, is_production),
              aes(quarter_end, point_estimate),
              colour = .palette$accent, linewidth = 1.0) +
    geom_point(data = dplyr::filter(bt, is_production),
               aes(quarter_end, point_estimate),
               colour = .palette$accent, size = 2.2) +
    facet_wrap(~ commodity, scales = "free_y",
               labeller = labeller(commodity = .commodity_labels)) +
    labs(x = NULL, y = "Quarterly exports (Mt)",
         title = "Walk-forward backtest: actuals vs candidate forecasts",
         subtitle = "Black: DISR actual. Blue: production-deployed spec. Grey: non-winning candidate specs (every backtest quarter, every spec).",
         caption = paste(
           "Each validation quarter refits every candidate on the",
           "expanding training window ending one quarter before, then",
           "predicts the held-out quarter. The production spec is the",
           "one with the lowest RMSE across this sample (see the",
           "diagnostics table above)."
         ))
}
```

## PortWatch tonnage trend

```{r tonnage-plot, fig.width=9, fig.height=3.5}
if (nrow(tonnage) > 0) {
  base <- tonnage |>
    mutate(quarter_end = as.Date(quarter_end),
           commodity   = factor(commodity, levels = com_levels)) |>
    filter(!is.na(commodity),
           quarter_end >= as.Date("2019-01-01"))

  # PortWatch coverage stops at max(obs_date), so the most-recent quarter
  # has a raw `tonnage` that under-represents the full quarter. Split:
  # plot completed quarters as the solid history line; show the partial
  # quarter as a hollow point at its extrapolated full-quarter value,
  # connected by a dashed segment.
  complete <- filter(base, is_complete)
  partial  <- filter(base, !is_complete, !is.na(tonnage_extrapolated))

  # Bridge segment from last-complete point to partial-extrapolated point
  bridge_seg <- if (nrow(complete) > 0 && nrow(partial) > 0) {
    complete |>
      group_by(commodity) |>
      filter(quarter_end == max(quarter_end)) |>
      ungroup() |>
      transmute(commodity,
                x = quarter_end, y = tonnage / 1e6) |>
      left_join(
        partial |> transmute(commodity,
                             xend = quarter_end,
                             yend = tonnage_extrapolated / 1e6),
        by = "commodity"
      ) |>
      filter(!is.na(.data$yend))
  } else {
    NULL
  }

  p <- ggplot() +
    geom_line(data = complete,
              aes(quarter_end, tonnage / 1e6, colour = commodity),
              linewidth = 0.6) +
    geom_point(data = complete,
               aes(quarter_end, tonnage / 1e6, colour = commodity),
               size = 1.4)
  if (!is.null(bridge_seg) && nrow(bridge_seg) > 0) {
    p <- p +
      geom_segment(data = bridge_seg,
                   aes(x = x, y = y, xend = xend, yend = yend,
                       colour = commodity),
                   linetype = "dashed", linewidth = 0.55)
  }
  if (nrow(partial) > 0) {
    p <- p +
      geom_point(data = partial,
                 aes(quarter_end, tonnage_extrapolated / 1e6,
                     colour = commodity),
                 size = 3.4, shape = 21, fill = .palette$bg, stroke = 1.4)
  }
  p +
    scale_colour_manual(values = .commodity_colours) +
    facet_wrap(~ commodity, scales = "free_y",
               labeller = labeller(commodity = .commodity_labels)) +
    labs(x = NULL, y = "Quarterly PortWatch tonnage (Mt)",
         title = "PortWatch quarterly tonnage by commodity",
         caption = paste(
           "Solid line: completed quarters (raw observed PortWatch sum).",
           "Hollow point: current partial quarter extrapolated to a",
           "full-quarter total using the same logic the bridge consumes."
         ))
}
```

## Change since last run

```{r delta, results='asis'}
# Reuse the `delta` tibble from the Top line; join back to nowcast to
# pull the now-point so this section stands alone.
if (nrow(nowcast_h0) == 0 || nrow(delta) == 0) {
  cat("*No nowcast available this run.*\n")
} else if (all(is.na(delta$delta_point))) {
  cat("*First run for every commodity (this quarter) -- no prior to compare against.*\n")
} else {
  nowcast_h0 |>
    select(commodity, quarter_end, now_Mt = point_estimate_Mt) |>
    left_join(delta, by = "commodity") |>
    filter(!is.na(delta_point)) |>
    mutate(
      commodity   = factor(commodity, levels = com_levels),
      prior_Mt    = round(now_Mt - delta_point, 2),
      now_Mt      = round(now_Mt, 2),
      delta_Mt    = round(delta_point, 2),
      delta_pct   = round(delta_pct, 2),
      hours_since = round(hours_since, 1)
    ) |>
    arrange(commodity) |>
    select(commodity, quarter_end, prior_Mt, now_Mt, delta_Mt, delta_pct,
           hours_since) |>
    knitr::kable(caption = "Same-quarter run-over-run delta. Drops rows for any commodity whose current quarter has no prior nowcast.")
}
```

## Flagged anomalies

```{r anomalies, results='asis'}
# Top 5 departures >2σ from the seasonal daily norm over the last 28
# days, sourced from `mart_latest_anomalies`. Port id resolves to a human
# name via `mart_dim_port` when available; we keep the raw id otherwise
# (synthetic fixtures use `portNNN` placeholders that have no metadata).
if (nrow(anomalies) == 0) {
  cat("*No departures >2σ from the seasonal norm in the last 28 days.*\n")
} else {
  port_name_lookup <- if (nrow(ports_dim) > 0 && "port_id" %in% names(ports_dim)) {
    stats::setNames(ports_dim$port_name, ports_dim$port_id)
  } else character()
  anomalies |>
    dplyr::arrange(dplyr::desc(abs(z_score))) |>
    dplyr::slice_head(n = 5) |>
    dplyr::mutate(
      port      = dplyr::coalesce(port_name_lookup[port_id], port_id),
      commodity = factor(commodity, levels = com_levels),
      `tonnage (kt)`  = round(tonnage  / 1e3, 0),
      `expected (kt)` = round(expected / 1e3, 0),
      `z`             = round(z_score, 2)
    ) |>
    dplyr::select(date = obs_date, port, commodity,
                  `tonnage (kt)`, `expected (kt)`, z) |>
    knitr::kable(
      caption = paste(
        "Largest |z|-score departures from the seasonal daily norm",
        "(±7-day window) in the last 28 days. Seasonal norm is built",
        "from training-window observations only; z > 0 is unusually",
        "busy, z < 0 unusually quiet."
      )
    )
}
```

## Methodology

```{r prod-equations-setup, include=FALSE}
fits_bench <- safe_read_wh("derived_bridge_fits_bench")
oos_csv    <- safe_read_csv(path(outputs, "bridge_diagnostics.csv"))

fmt_signed <- function(x, digits = 3) {
  if (!is.finite(x)) return("")
  paste0(if (x >= 0) "+ " else "- ", formatC(abs(x), digits = digits,
                                              format = "f"))
}

# Build a per-commodity entry with the production equation string and a
# one-line "why" sentence drawing on the live OOS leaderboard.
build_eq_entry <- function(com) {
  prod_row <- oos_csv |>
    filter(commodity == com,
           isTRUE(production_choice) | production_choice %in% c(TRUE,"TRUE","true")) |>
    slice(1)
  if (nrow(prod_row) == 0) return(NULL)

  spec <- prod_row$spec[1]

  oos_c <- oos_csv |> filter(commodity == com) |> arrange(rmse_valid)
  best  <- oos_c$rmse_valid[1]
  naive <- oos_c$rmse_naive[1]
  ratio <- best / naive

  # Equation TeX + a Wald-pvalue narrative pulled from the same row.
  is_single <- length(fits_bench) > 0 &&
               paste0(com, "__", spec) %in% names(fits_bench)
  fit_key   <- paste0(com, "__", spec)

  if (is_single) {
    co <- stats::coef(fits_bench[[fit_key]]$fit)
    if (spec == "aggregate") {
      eq <- sprintf(
        "\\log V_{Q} \\;=\\; %.3f \\;%s \\,\\Delta_{4}\\log T_{Q} \\;%s \\,\\log V_{Q-4} \\;+\\; \\varepsilon_{Q}",
        co[["(Intercept)"]],
        fmt_signed(co[["yoy_log_tonnage"]], 3),
        fmt_signed(co[["log_volume_lag4"]], 3)
      )
    } else if (spec == "midas") {
      eq <- sprintf(
        "\\log V_{Q} \\;=\\; %.3f \\;%s \\,\\Delta_{4}\\log T_{Q,m1} \\;%s \\,\\Delta_{4}\\log T_{Q,m2} \\;%s \\,\\Delta_{4}\\log T_{Q,m3} \\;%s \\,\\log V_{Q-4} \\;+\\; \\varepsilon_{Q}",
        co[["(Intercept)"]],
        fmt_signed(co[["yoy_log_tonnage_m1"]], 3),
        fmt_signed(co[["yoy_log_tonnage_m2"]], 3),
        fmt_signed(co[["yoy_log_tonnage_m3"]], 3),
        fmt_signed(co[["log_volume_lag4"]], 3)
      )
    } else if (spec == "bojo") {
      eq <- sprintf(
        "\\Delta_{4}\\log V_{Q} \\;=\\; %.3f \\;%s \\,\\Delta_{4}\\log T_{Q} \\;+\\; \\varepsilon_{Q}",
        co[["(Intercept)"]],
        fmt_signed(co[["yoy_log_tonnage"]], 3)
      )
    } else if (spec == "lagged") {
      eq <- sprintf(
        "\\log V_{Q} \\;=\\; %.3f \\;%s \\,\\Delta_{4}\\log T_{Q} \\;%s \\,\\Delta_{4}\\log T_{Q-1} \\;%s \\,\\log V_{Q-4} \\;+\\; \\varepsilon_{Q}",
        co[["(Intercept)"]],
        fmt_signed(co[["yoy_log_tonnage"]], 3),
        fmt_signed(co[["yoy_log_tonnage_lag1"]], 3),
        fmt_signed(co[["log_volume_lag4"]], 3)
      )
    } else if (spec == "price_aug") {
      eq <- sprintf(
        "\\log V_{Q} \\;=\\; %.3f \\;%s \\,\\Delta_{4}\\log T_{Q} \\;%s \\,\\Delta_{4}\\log P_{Q} \\;%s \\,\\log V_{Q-4} \\;+\\; \\varepsilon_{Q}",
        co[["(Intercept)"]],
        fmt_signed(co[["yoy_log_tonnage"]], 3),
        fmt_signed(co[["yoy_log_price"]], 3),
        fmt_signed(co[["log_volume_lag4"]], 3)
      )
    } else if (spec == "demand_aug") {
      # demand_aug's RHS varies per commodity (iron_ore: CLI, coal_thermal:
      # exports). Render the surviving demand regressor(s) as a sum.
      demand_terms <- grep("^yoy_log_demand_", names(co), value = TRUE)
      demand_tex <- paste(
        vapply(demand_terms, function(t) {
          sprintf("%s \\,\\Delta_{4}\\log D^{\\,%s}_{Q}",
                  fmt_signed(co[[t]], 3),
                  sub("^yoy_log_demand_", "", t))
        }, character(1)),
        collapse = " \\;"
      )
      eq <- sprintf(
        "\\log V_{Q} \\;=\\; %.3f \\;%s \\,\\Delta_{4}\\log T_{Q} \\;%s \\;%s \\,\\log V_{Q-4} \\;+\\; \\varepsilon_{Q}",
        co[["(Intercept)"]],
        fmt_signed(co[["yoy_log_tonnage"]], 3),
        demand_tex,
        fmt_signed(co[["log_volume_lag4"]], 3)
      )
    } else {
      eq <- ""
    }
  } else {
    # Combination winner: show the weighted sum across components.
    comps <- oos_c |> filter(spec %in% c("aggregate","midas","bojo")) |>
                       arrange(rmse_valid)
    eq <- sprintf(
      "\\hat V_{Q}^{\\,%s} \\;=\\; \\sum_{i \\in \\{%s\\}} w_i \\hat V_{Q}^{\\,(i)}",
      spec, paste(comps$spec, collapse = ", ")
    )
  }

  # Lock4 Wald p sentence -- pull from this commodity's *single* specs.
  agg_p <- oos_c |> filter(spec == "aggregate") |> dplyr::pull(beta_lag4_eq_1_pval)
  mid_p <- oos_c |> filter(spec == "midas")     |> dplyr::pull(beta_lag4_eq_1_pval)
  wald_line <- if (length(agg_p) && length(mid_p) &&
                   is.finite(agg_p[1]) && is.finite(mid_p[1])) {
    sprintf("Wald test of $\\beta_{lag4}=1$: aggregate p = %.2f, midas p = %.2f.",
            agg_p[1], mid_p[1])
  } else { "" }

  rationale <- if (spec == "bojo") {
    sprintf("%s Restriction not rejected; pure YoY-on-YoY wins the OOS leaderboard at RMSE = %.2f Mt (%.2f× naive).",
            wald_line, best, ratio)
  } else if (spec == "aggregate") {
    sprintf("%s Restriction rejected; the free seasonal-anchor coefficient is doing real work (β₄ ≈ %.2f). OOS RMSE = %.2f Mt (%.2f× naive).",
            wald_line,
            unname(stats::coef(fits_bench[[fit_key]]$fit)["log_volume_lag4"]),
            best, ratio)
  } else if (spec == "midas") {
    sprintf("%s OOS RMSE = %.2f Mt (%.2f× naive); within-quarter monthly timing carries enough signal to justify the extra parameters.",
            wald_line, best, ratio)
  } else {
    sprintf("Combination of single-spec forecasts; OOS RMSE = %.2f Mt (%.2f× naive).",
            best, ratio)
  }

  list(spec = spec, eq = eq, rationale = rationale)
}

prod_entries <- lapply(unique(nowcast_h0$commodity), function(com) {
  list(com = com, entry = build_eq_entry(com))
})
```

### Production equation per commodity

These are the actual equations driving the headline nowcast above. One per commodity, picked by walk-forward backtest RMSE from a bench of `{aggregate, midas, bojo, lagged, price_aug, demand_aug, equal_avg, inv_mse}` specs.

```{r prod-equations-show, results='asis'}
emit <- function(label, entry) {
  if (is.null(entry)) return(invisible())
  cat(sprintf("\n#### %s — `%s` spec\n\n", label, entry$spec))
  cat(sprintf("$$%s$$\n\n", entry$eq))
  cat(entry$rationale, "\n\n")
}
for (pe in prod_entries) emit(commodity_label(pe$com), pe$entry)
```

### Conventions

- $V_{Q}$ is DISR Resources & Energy Quarterly Table 16 physical tonnage (Mt) for the commodity. Coal is metallurgical + thermal summed.
- $T_{Q}$ is PortWatch AIS daily tonnage summed across the commodity's ports over quarter $Q$. The current-quarter partial-month scale-up anchors on the data's latest observation date (not `Sys.Date()`), so PortWatch's 7–14 day refresh lag doesn't bias the estimate down.
- $\Delta_{4}$ is the year-on-year log difference. The YoY transform strips deterministic seasonality, so no explicit STL / X-13 step is needed at our sample size (n ≈ 23 quarters).
- Newey-West HAC standard errors with `lag = 1`. Uncertainty bands from residual bootstrap, scaled by $\sqrt{1 - \text{share\_observed}}$ so they collapse as the current quarter fills in.
- Full derivation, candidate bench, and the literature this draws on: [`docs/METHODOLOGY.md`](https://github.com/DavidAStephan/ResourceExplorer/blob/main/docs/METHODOLOGY.md).
