---
title: "Australian Resource Export Volumes"
subtitle: "Quarterly nowcast — iron ore and coal"
date: "`r format(Sys.Date(), '%e %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    self_contained: true
    theme: null
    highlight: null
    css: style.css
    mathjax: default
    md_extensions: "+raw_html"
params:
  outputs_dir: "../../outputs"
  warehouse_dir: "../../data/warehouse"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                     dev = "png", dpi = 144)
library(dplyr)
library(readr)
library(ggplot2)
library(fs)

# Unified palette & ggplot theme so charts feel like part of the page,
# not a foreign bitmap.
.palette <- list(
  text       = "#1A1F2C",
  text_muted = "#5A6478",
  border     = "#E4E1D9",
  accent     = "#1C3D5A",
  accent2    = "#8A4A1F",
  grid       = "#EFEDE7",
  bg         = "#FAFAF7"
)
theme_rt <- function(base_size = 11) {
  theme_minimal(base_family = "Inter", base_size = base_size) +
    theme(
      plot.background  = element_rect(fill = .palette$bg, colour = NA),
      panel.background = element_rect(fill = .palette$bg, colour = NA),
      panel.grid.major = element_line(colour = .palette$grid, size = 0.4),
      panel.grid.minor = element_blank(),
      axis.text  = element_text(colour = .palette$text_muted),
      axis.title = element_text(colour = .palette$text_muted, size = base_size - 1),
      axis.line.x = element_line(colour = .palette$border, size = 0.5),
      axis.ticks  = element_line(colour = .palette$border, size = 0.4),
      strip.text  = element_text(colour = .palette$text, face = "bold",
                                 hjust = 0, size = base_size + 1),
      strip.background = element_blank(),
      plot.title = element_text(colour = .palette$text, face = "bold",
                                size = base_size + 2,
                                margin = margin(0, 0, 8, 0)),
      plot.subtitle = element_text(colour = .palette$text_muted,
                                   size = base_size - 1,
                                   margin = margin(0, 0, 12, 0)),
      plot.caption  = element_text(colour = .palette$text_muted,
                                   size = base_size - 2, hjust = 0,
                                   margin = margin(12, 0, 0, 0)),
      legend.position = "none"
    )
}
theme_set(theme_rt())
.commodity_colours <- c(iron_ore = .palette$accent,
                         coal     = .palette$accent2)

outputs <- params$outputs_dir
wh_dir  <- params$warehouse_dir

safe_read_csv <- function(p) {
  if (!file_exists(p)) return(tibble())
  suppressMessages(read_csv(p, show_col_types = FALSE))
}
safe_read_wh <- function(name) {
  p <- path(wh_dir, paste0(name, ".rds"))
  if (!file_exists(p)) return(tibble())
  tibble::as_tibble(readRDS(p))
}

nowcast  <- safe_read_csv(path(outputs, "nowcast_current.csv"))
tonnage  <- safe_read_csv(path(outputs, "tonnage_quarterly.csv"))
diag_fit <- safe_read_csv(path(outputs, "bridge_diagnostics.csv"))
history  <- safe_read_wh("mart_nowcast_history")
disr     <- safe_read_wh("raw_disr_req_quarterly")

com_levels <- c("iron_ore", "coal")

# For each commodity in `nowcast`, find the most-recent prior row in
# `history` for the same quarter that isn't this run, and compute the
# point + 80% band delta. Returns a tibble with one row per commodity;
# delta cols are NA when no prior exists.
compute_wow_delta <- function(nowcast, history) {
  if (nrow(nowcast) == 0 || nrow(history) == 0) {
    return(tibble::tibble(commodity = character(),
                          delta_point = double(),
                          delta_pct = double(),
                          prior_run = as.POSIXct(NA),
                          hours_since = double()))
  }
  history <- history |>
    mutate(run_timestamp = as.POSIXct(run_timestamp),
           quarter_end   = as.Date(quarter_end))
  purrr::map_dfr(seq_len(nrow(nowcast)), function(i) {
    cur <- nowcast[i, ]
    cur_ts <- as.POSIXct(cur$run_timestamp)
    prior <- history |>
      filter(commodity == cur$commodity,
             quarter_end == as.Date(cur$quarter_end),
             run_timestamp < cur_ts) |>
      arrange(desc(run_timestamp)) |>
      slice(1)
    if (nrow(prior) == 0) {
      tibble::tibble(commodity   = cur$commodity,
                     delta_point = NA_real_,
                     delta_pct   = NA_real_,
                     prior_run   = as.POSIXct(NA),
                     hours_since = NA_real_)
    } else {
      d <- cur$point_estimate_Mt - prior$point_estimate
      tibble::tibble(
        commodity   = cur$commodity,
        delta_point = d,
        delta_pct   = 100 * d / prior$point_estimate,
        prior_run   = prior$run_timestamp,
        hours_since = as.numeric(difftime(cur_ts, prior$run_timestamp,
                                          units = "hours"))
      )
    }
  })
}

delta <- compute_wow_delta(nowcast, history)
```

## Headline nowcasts

```{r top-line, results='asis'}
commodity_label <- function(x) {
  switch(x,
         iron_ore = "Iron ore",
         coal     = "Coal",
         tools::toTitleCase(gsub("_", " ", x)))
}

if (nrow(nowcast) == 0) {
  cat('<div class="callout"><span class="label">Status</span>',
      'No nowcast produced this run.</div>')
} else {
  nc <- nowcast |>
    mutate(commodity = factor(commodity, levels = com_levels)) |>
    arrange(commodity) |>
    left_join(delta, by = "commodity")
  share_pct <- round(100 * nc$share_observed[1])
  qtr_label <- format(as.Date(nc$quarter_end[1]), "%B %Y")

  cat(sprintf(
    '<p class="kicker">Current quarter — %s · %d%% of the quarter observed</p>\n',
    qtr_label, share_pct
  ))
  cat('<div class="headline-grid">\n')
  for (i in seq_len(nrow(nc))) {
    r <- nc[i, ]
    delta_class <- if (is.na(r$delta_point)) "flat"
                   else if (r$delta_point >  0.05) "up"
                   else if (r$delta_point < -0.05) "down"
                   else "flat"
    delta_text <- if (is.na(r$delta_point)) {
      "First nowcast for this quarter"
    } else {
      sign_lbl  <- if (r$delta_point >= 0) "+" else "−"
      sign_pct  <- if (r$delta_pct   >= 0) "+" else "−"
      hours_lbl <- if (r$hours_since < 24) sprintf("%.0fh ago", r$hours_since)
                   else sprintf("%.0f days ago", r$hours_since / 24)
      sprintf("%s%.1f Mt&nbsp;&nbsp;(%s%.1f%%) vs the run %s",
              sign_lbl, abs(r$delta_point),
              sign_pct, abs(r$delta_pct),
              hours_lbl)
    }
    # Single-line HTML so pandoc treats the whole card as one raw-HTML
    # block and doesn't paragraph-wrap the inline spans.
    card_html <- sprintf(
      '<div class="headline-card"><span class="label">%s</span><div class="value-row"><span class="value">%.1f</span><span class="unit">Mt</span></div><span class="ci">80%% CI · %.1f – %.1f Mt</span><span class="delta %s">%s</span></div>',
      commodity_label(as.character(r$commodity)),
      r$point_estimate_Mt,
      r$lower_80, r$upper_80,
      delta_class, delta_text
    )
    cat(card_html, "\n", sep = "")
  }
  cat('</div>\n')
}
```

## Nowcast vs recent actuals

```{r actuals-plot, fig.width=9, fig.height=4.5}
if (nrow(disr) > 0 && nrow(nowcast) > 0) {
  hist <- disr |>
    mutate(quarter_end = as.Date(quarter_end),
           commodity   = factor(commodity, levels = com_levels)) |>
    filter(!is.na(commodity),
           quarter_end >= as.Date("2019-01-01")) |>
    arrange(commodity, quarter_end)

  nc <- nowcast |>
    mutate(quarter_end = as.Date(quarter_end),
           commodity   = factor(commodity, levels = com_levels))

  ggplot(hist, aes(quarter_end, tonnes_Mt)) +
    geom_line(colour = .palette$text_muted, linewidth = 0.5) +
    geom_point(colour = .palette$text_muted, size = 1.2) +
    geom_linerange(data = nc,
                   aes(x = quarter_end,
                       ymin = lower_95, ymax = upper_95,
                       y = point_estimate_Mt),
                   colour = .palette$accent, alpha = 0.25, linewidth = 1.4) +
    geom_linerange(data = nc,
                   aes(x = quarter_end,
                       ymin = lower_80, ymax = upper_80,
                       y = point_estimate_Mt),
                   colour = .palette$accent, alpha = 0.55, linewidth = 2.2) +
    geom_point(data = nc,
               aes(quarter_end, point_estimate_Mt),
               colour = .palette$accent, size = 3.4) +
    facet_wrap(~ commodity, scales = "free_y",
               labeller = labeller(commodity = c(iron_ore = "Iron ore",
                                                  coal     = "Coal"))) +
    labs(x = NULL, y = "Quarterly exports (Mt)",
         title = "DISR quarterly export volume, with current-quarter nowcast",
         subtitle = "Grey: published DISR Resources & Energy Quarterly Table 16 actuals. Blue: current-quarter nowcast with 80% (dark) and 95% (light) intervals.")
}
```

## Bridge diagnostics

```{r diagnostics}
if (nrow(diag_fit) > 0) {
  diag_fit |>
    mutate(
      commodity      = factor(commodity, levels = com_levels),
      production     = ifelse(isTRUE(production_choice) |
                              production_choice %in% c("TRUE","true"),
                              "★", ""),
      beta_tonnage   = ifelse(is.na(beta_tonnage), "",
                              sprintf("%.3f (se %.3f)",
                                      beta_tonnage, beta_tonnage_se)),
      beta_lag4      = ifelse(is.na(beta_lag4), "",
                              sprintf("%.3f", beta_lag4)),
      lag4_eq_1_pval = ifelse(is.na(beta_lag4_eq_1_pval), "",
                              sprintf("%.3f", beta_lag4_eq_1_pval)),
      r2_train       = ifelse(is.na(r_squared), "",
                              sprintf("%.2f", r_squared)),
      r2_oos         = ifelse(is.na(r_squared_oos), "",
                              sprintf("%.2f", r_squared_oos)),
      rmse_valid     = ifelse(is.na(rmse_valid), "",
                              sprintf("%.2f", rmse_valid)),
      ratio_vs_naive = ifelse(is.na(ratio_vs_naive), "",
                              sprintf("%.2f", ratio_vs_naive))
    ) |>
    arrange(commodity, spec) |>
    select(`★` = production,
           commodity, spec,
           `R²` = r2_train, `R²_oos` = r2_oos,
           `β_T (se)` = beta_tonnage,
           `β_lag4` = beta_lag4,
           `p(β_lag4=1)` = lag4_eq_1_pval,
           rmse_valid,
           `RMSE/naive` = ratio_vs_naive) |>
    knitr::kable(
      caption = paste(
        "★ marks the production-deployed spec per commodity (best of the",
        "candidate set + combinations by backtest RMSE). `R²` is",
        "in-sample, `R²_oos` is walk-forward out-of-sample.",
        "`p(β_lag4=1)` is the Wald-test p-value for the YoY-on-YoY",
        "restriction (NA for `bojo`, which imposes it; `equal_avg` and",
        "`inv_mse` are combination forecasts and have no in-sample β).",
        "`RMSE/naive` < 1 means the spec beats the seasonal random walk."
      )
    )
}
```

## PortWatch tonnage trend

```{r tonnage-plot, fig.width=9, fig.height=3.5}
if (nrow(tonnage) > 0) {
  base <- tonnage |>
    mutate(quarter_end = as.Date(quarter_end),
           commodity   = factor(commodity, levels = com_levels)) |>
    filter(!is.na(commodity),
           quarter_end >= as.Date("2019-01-01"))

  # PortWatch coverage stops at max(obs_date), so the most-recent quarter
  # has a raw `tonnage` that under-represents the full quarter. Split:
  # plot completed quarters as the solid history line; show the partial
  # quarter as a hollow point at its extrapolated full-quarter value,
  # connected by a dashed segment.
  complete <- filter(base, is_complete)
  partial  <- filter(base, !is_complete, !is.na(tonnage_extrapolated))

  # Bridge segment from last-complete point to partial-extrapolated point
  bridge_seg <- if (nrow(complete) > 0 && nrow(partial) > 0) {
    complete |>
      group_by(commodity) |>
      filter(quarter_end == max(quarter_end)) |>
      ungroup() |>
      transmute(commodity,
                x = quarter_end, y = tonnage / 1e6) |>
      left_join(
        partial |> transmute(commodity,
                             xend = quarter_end,
                             yend = tonnage_extrapolated / 1e6),
        by = "commodity"
      ) |>
      filter(!is.na(.data$yend))
  } else {
    NULL
  }

  p <- ggplot() +
    geom_line(data = complete,
              aes(quarter_end, tonnage / 1e6, colour = commodity),
              linewidth = 0.6) +
    geom_point(data = complete,
               aes(quarter_end, tonnage / 1e6, colour = commodity),
               size = 1.4)
  if (!is.null(bridge_seg) && nrow(bridge_seg) > 0) {
    p <- p +
      geom_segment(data = bridge_seg,
                   aes(x = x, y = y, xend = xend, yend = yend,
                       colour = commodity),
                   linetype = "dashed", linewidth = 0.55)
  }
  if (nrow(partial) > 0) {
    p <- p +
      geom_point(data = partial,
                 aes(quarter_end, tonnage_extrapolated / 1e6,
                     colour = commodity),
                 size = 3.4, shape = 21, fill = .palette$bg, stroke = 1.4)
  }
  p +
    scale_colour_manual(values = .commodity_colours) +
    facet_wrap(~ commodity, scales = "free_y",
               labeller = labeller(commodity = c(iron_ore = "Iron ore",
                                                  coal     = "Coal"))) +
    labs(x = NULL, y = "Quarterly PortWatch tonnage (Mt)",
         title = "PortWatch quarterly tonnage by commodity",
         caption = paste(
           "Solid line: completed quarters (raw observed PortWatch sum).",
           "Hollow point: current partial quarter extrapolated to a",
           "full-quarter total using the same logic the bridge consumes."
         ))
}
```

## Change since last run

```{r delta, results='asis'}
# Reuse the `delta` tibble from the Top line; join back to nowcast to
# pull the now-point so this section stands alone.
if (nrow(nowcast) == 0 || nrow(delta) == 0) {
  cat("*No nowcast available this run.*\n")
} else if (all(is.na(delta$delta_point))) {
  cat("*First run for every commodity (this quarter) -- no prior to compare against.*\n")
} else {
  nowcast |>
    select(commodity, quarter_end, now_Mt = point_estimate_Mt) |>
    left_join(delta, by = "commodity") |>
    filter(!is.na(delta_point)) |>
    mutate(
      commodity   = factor(commodity, levels = com_levels),
      prior_Mt    = round(now_Mt - delta_point, 2),
      now_Mt      = round(now_Mt, 2),
      delta_Mt    = round(delta_point, 2),
      delta_pct   = round(delta_pct, 2),
      hours_since = round(hours_since, 1)
    ) |>
    arrange(commodity) |>
    select(commodity, quarter_end, prior_Mt, now_Mt, delta_Mt, delta_pct,
           hours_since) |>
    knitr::kable(caption = "Same-quarter run-over-run delta. Drops rows for any commodity whose current quarter has no prior nowcast.")
}
```

## Methodology

```{r prod-equations-setup, include=FALSE}
fits_bench <- safe_read_wh("derived_bridge_fits_bench")
oos_csv    <- safe_read_csv(path(outputs, "bridge_diagnostics.csv"))

fmt_signed <- function(x, digits = 3) {
  if (!is.finite(x)) return("")
  paste0(if (x >= 0) "+ " else "- ", formatC(abs(x), digits = digits,
                                              format = "f"))
}

# Build a per-commodity entry with the production equation string and a
# one-line "why" sentence drawing on the live OOS leaderboard.
build_eq_entry <- function(com) {
  prod_row <- oos_csv |>
    filter(commodity == com,
           isTRUE(production_choice) | production_choice %in% c(TRUE,"TRUE","true")) |>
    slice(1)
  if (nrow(prod_row) == 0) return(NULL)

  spec <- prod_row$spec[1]

  oos_c <- oos_csv |> filter(commodity == com) |> arrange(rmse_valid)
  best  <- oos_c$rmse_valid[1]
  naive <- oos_c$rmse_naive[1]
  ratio <- best / naive

  # Equation TeX + a Wald-pvalue narrative pulled from the same row.
  is_single <- length(fits_bench) > 0 &&
               paste0(com, "__", spec) %in% names(fits_bench)
  fit_key   <- paste0(com, "__", spec)

  if (is_single) {
    co <- stats::coef(fits_bench[[fit_key]]$fit)
    if (spec == "aggregate") {
      eq <- sprintf(
        "\\log V_{Q} \\;=\\; %.3f \\;%s \\,\\Delta_{4}\\log T_{Q} \\;%s \\,\\log V_{Q-4} \\;+\\; \\varepsilon_{Q}",
        co[["(Intercept)"]],
        fmt_signed(co[["yoy_log_tonnage"]], 3),
        fmt_signed(co[["log_volume_lag4"]], 3)
      )
    } else if (spec == "midas") {
      eq <- sprintf(
        "\\log V_{Q} \\;=\\; %.3f \\;%s \\,\\Delta_{4}\\log T_{Q,m1} \\;%s \\,\\Delta_{4}\\log T_{Q,m2} \\;%s \\,\\Delta_{4}\\log T_{Q,m3} \\;%s \\,\\log V_{Q-4} \\;+\\; \\varepsilon_{Q}",
        co[["(Intercept)"]],
        fmt_signed(co[["yoy_log_tonnage_m1"]], 3),
        fmt_signed(co[["yoy_log_tonnage_m2"]], 3),
        fmt_signed(co[["yoy_log_tonnage_m3"]], 3),
        fmt_signed(co[["log_volume_lag4"]], 3)
      )
    } else if (spec == "bojo") {
      eq <- sprintf(
        "\\Delta_{4}\\log V_{Q} \\;=\\; %.3f \\;%s \\,\\Delta_{4}\\log T_{Q} \\;+\\; \\varepsilon_{Q}",
        co[["(Intercept)"]],
        fmt_signed(co[["yoy_log_tonnage"]], 3)
      )
    } else {
      eq <- ""
    }
  } else {
    # Combination winner: show the weighted sum across components.
    comps <- oos_c |> filter(spec %in% c("aggregate","midas","bojo")) |>
                       arrange(rmse_valid)
    eq <- sprintf(
      "\\hat V_{Q}^{\\,%s} \\;=\\; \\sum_{i \\in \\{%s\\}} w_i \\hat V_{Q}^{\\,(i)}",
      spec, paste(comps$spec, collapse = ", ")
    )
  }

  # Lock4 Wald p sentence -- pull from this commodity's *single* specs.
  agg_p <- oos_c |> filter(spec == "aggregate") |> dplyr::pull(beta_lag4_eq_1_pval)
  mid_p <- oos_c |> filter(spec == "midas")     |> dplyr::pull(beta_lag4_eq_1_pval)
  wald_line <- if (length(agg_p) && length(mid_p) &&
                   is.finite(agg_p[1]) && is.finite(mid_p[1])) {
    sprintf("Wald test of $\\beta_{lag4}=1$: aggregate p = %.2f, midas p = %.2f.",
            agg_p[1], mid_p[1])
  } else { "" }

  rationale <- if (spec == "bojo") {
    sprintf("%s Restriction not rejected; pure YoY-on-YoY wins the OOS leaderboard at RMSE = %.2f Mt (%.2f× naive).",
            wald_line, best, ratio)
  } else if (spec == "aggregate") {
    sprintf("%s Restriction rejected; the free seasonal-anchor coefficient is doing real work (β₄ ≈ %.2f). OOS RMSE = %.2f Mt (%.2f× naive).",
            wald_line,
            unname(stats::coef(fits_bench[[fit_key]]$fit)["log_volume_lag4"]),
            best, ratio)
  } else if (spec == "midas") {
    sprintf("%s OOS RMSE = %.2f Mt (%.2f× naive); within-quarter monthly timing carries enough signal to justify the extra parameters.",
            wald_line, best, ratio)
  } else {
    sprintf("Combination of single-spec forecasts; OOS RMSE = %.2f Mt (%.2f× naive).",
            best, ratio)
  }

  list(spec = spec, eq = eq, rationale = rationale)
}

iron_eq <- build_eq_entry("iron_ore")
coal_eq <- build_eq_entry("coal")
```

### Production equation per commodity

These are the actual equations driving the headline nowcast above. They're picked per commodity by walk-forward backtest RMSE from a bench of `{aggregate, midas, bojo, equal_avg, inv_mse}` specs.

```{r prod-equations-show, results='asis'}
emit <- function(label, entry) {
  if (is.null(entry)) return(invisible())
  cat(sprintf("\n#### %s — `%s` spec\n\n", label, entry$spec))
  cat(sprintf("$$%s$$\n\n", entry$eq))
  cat(entry$rationale, "\n\n")
}
emit("Iron ore", iron_eq)
emit("Coal",     coal_eq)
```

### Conventions

- $V_{Q}$ is DISR Resources & Energy Quarterly Table 16 physical tonnage (Mt) for the commodity. Coal is metallurgical + thermal summed.
- $T_{Q}$ is PortWatch AIS daily tonnage summed across the commodity's ports over quarter $Q$. The current-quarter partial-month scale-up anchors on the data's latest observation date (not `Sys.Date()`), so PortWatch's 7–14 day refresh lag doesn't bias the estimate down.
- $\Delta_{4}$ is the year-on-year log difference. The YoY transform strips deterministic seasonality, so no explicit STL / X-13 step is needed at our sample size (n ≈ 23 quarters).
- Newey-West HAC standard errors with `lag = 1`. Uncertainty bands from residual bootstrap, scaled by $\sqrt{1 - \text{share\_observed}}$ so they collapse as the current quarter fills in.
- Full derivation, candidate bench, and the literature this draws on: [`docs/METHODOLOGY.md`](https://github.com/DavidAStephan/ResourceExplorer/blob/main/docs/METHODOLOGY.md).
