---
title: "Australian Real Goods Nowcast — Briefing"
subtitle: "resourcetracker · weekly run"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    self_contained: true
    theme: cosmo
params:
  outputs_dir: "../../outputs"
  warehouse_dir: "../../data/warehouse"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(readr)
library(ggplot2)
library(fs)

outputs <- params$outputs_dir
wh_dir  <- params$warehouse_dir

safe_read_csv <- function(p) {
  if (!file_exists(p)) return(tibble())
  suppressMessages(read_csv(p, show_col_types = FALSE))
}

safe_read_wh <- function(name) {
  p <- path(wh_dir, paste0(name, ".rds"))
  if (!file_exists(p)) return(tibble())
  tibble::as_tibble(readRDS(p))
}

nowcast  <- safe_read_csv(path(outputs, "nowcast_current.csv"))
tonnage  <- safe_read_csv(path(outputs, "tonnage_monthly.csv"))
diag_fit <- safe_read_csv(path(outputs, "bridge_diagnostics.csv"))

history  <- safe_read_wh("mart_nowcast_history")
abs_5302 <- safe_read_wh("raw_abs_5302_quarterly")
anoms    <- safe_read_wh("mart_latest_anomalies")
```

## Headline nowcast

```{r headline, results='asis'}
if (nrow(nowcast) == 0 || is.na(nowcast$point_estimate[1])) {
  cat("*No nowcast available this run.*")
} else {
  nc <- nowcast[1, ]
  cat(sprintf(
    "**Quarter %s:** A$%s million (80%% CI: A$%s – A$%s; 95%% CI: A$%s – A$%s). Share of quarter observed: %.0f%%.",
    format(nc$quarter_end, "%b %Y"),
    formatC(nc$point_estimate, format = "d", big.mark = ","),
    formatC(nc$lower_80,       format = "d", big.mark = ","),
    formatC(nc$upper_80,       format = "d", big.mark = ","),
    formatC(nc$lower_95,       format = "d", big.mark = ","),
    formatC(nc$upper_95,       format = "d", big.mark = ","),
    100 * nc$share_observed
  ))
}
```

```{r headline-plot, fig.width=7, fig.height=3.5}
if (nrow(history) > 0) {
  history |>
    mutate(quarter_end = as.Date(quarter_end)) |>
    ggplot(aes(run_timestamp, point_estimate)) +
    geom_ribbon(aes(ymin = lower_95, ymax = upper_95), alpha = 0.15) +
    geom_ribbon(aes(ymin = lower_80, ymax = upper_80), alpha = 0.25) +
    geom_line() +
    labs(x = "Run timestamp",
         y = "Real goods credits (A$m, chain-volume)",
         title = "Nowcast evolution within the current quarter") +
    theme_minimal()
}
```

## Commodity decomposition

```{r decomposition}
if (nrow(tonnage) > 0) {
  tonnage |>
    filter(!is.na(commodity)) |>
    group_by(commodity) |>
    summarise(tonnage = sum(tonnage, na.rm = TRUE), .groups = "drop") |>
    mutate(share = tonnage / sum(tonnage)) |>
    arrange(desc(share)) |>
    knitr::kable(digits = c(NA, 0, 3),
                 col.names = c("Commodity", "Tonnage (summed)", "Share"))
} else {
  cat("*Tonnage panel empty.*")
}
```

## Change since last run

```{r delta}
if (nrow(history) >= 2) {
  ordered <- history[order(history$run_timestamp), ]
  current <- ordered[nrow(ordered), ]
  prev    <- ordered[nrow(ordered) - 1, ]
  tibble(
    metric = c("Point estimate", "80% lower", "80% upper",
               "95% lower", "95% upper", "Share observed"),
    prior  = c(prev$point_estimate, prev$lower_80, prev$upper_80,
               prev$lower_95, prev$upper_95, prev$share_observed),
    now    = c(current$point_estimate, current$lower_80, current$upper_80,
               current$lower_95, current$upper_95, current$share_observed),
    delta  = c(current$point_estimate - prev$point_estimate,
               current$lower_80       - prev$lower_80,
               current$upper_80       - prev$upper_80,
               current$lower_95       - prev$lower_95,
               current$upper_95       - prev$upper_95,
               current$share_observed - prev$share_observed)
  ) |>
    knitr::kable(digits = c(NA, 1, 1, 1))
} else {
  cat("*First run this quarter -- no prior run to compare against.*")
}
```

## Comparison to ABS latest actual

```{r abs-compare, results='asis'}
if (nrow(abs_5302) > 0) {
  latest_abs <- abs_5302 |>
    filter(!is.na(value_chainvol_aud_m), value_chainvol_aud_m > 0) |>
    arrange(desc(quarter_end)) |>
    slice(1)
  cat(sprintf(
    "Latest ABS 5302.0 actual (quarter ending %s): **A$%s million**.\n\n",
    format(as.Date(latest_abs$quarter_end), "%b %Y"),
    formatC(latest_abs$value_chainvol_aud_m,
            format = "d", big.mark = ",")
  ))
} else {
  cat("*ABS 5302.0 panel empty.*")
}
```

## Flagged anomalies

```{r anomalies-table}
if (nrow(anoms) > 0) {
  anoms |>
    arrange(desc(abs(z_score))) |>
    slice_head(n = 10) |>
    select(obs_date, port_id, commodity, tonnage, z_score) |>
    knitr::kable(digits = c(NA, NA, NA, 0, 2))
} else {
  cat("*No >2σ departures flagged in the last 28 days.*")
}
```

## Methodology footnote

Bridge regression per commodity on log levels with an AR(1) lag:

$$\log y_{c,m} = \beta_0 + \beta_1 \log T^{SA}_{c,m} + \beta_2 \log P_{c,m} + \beta_3 \log y_{c,m-1} + \varepsilon_{c,m}$$

Fit on 2019-01 → 2023-12; validated on 2024+. Newey-West (Bartlett) SEs
with lag 3. Seasonal adjustment via `stats::stl()` (periodic, robust).
Quarterly aggregate converted to chain-volume via the ABS 5302.0
implicit deflator (trailing 4-quarter mean as forecast). Uncertainty
bands via 1,000 bootstrap draws of bridge residuals, scaled by
$\sqrt{1 - \text{share\_observed}}$.
