---
title: "Australian Resource Export Volumes — Quarterly Nowcast"
subtitle: "resourcetracker · per-commodity tonnage nowcast (iron_ore, coal)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    self_contained: true
    theme: cosmo
params:
  outputs_dir: "../../outputs"
  warehouse_dir: "../../data/warehouse"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(readr)
library(ggplot2)
library(fs)

outputs <- params$outputs_dir
wh_dir  <- params$warehouse_dir

safe_read_csv <- function(p) {
  if (!file_exists(p)) return(tibble())
  suppressMessages(read_csv(p, show_col_types = FALSE))
}
safe_read_wh <- function(name) {
  p <- path(wh_dir, paste0(name, ".rds"))
  if (!file_exists(p)) return(tibble())
  tibble::as_tibble(readRDS(p))
}

nowcast  <- safe_read_csv(path(outputs, "nowcast_current.csv"))
tonnage  <- safe_read_csv(path(outputs, "tonnage_quarterly.csv"))
diag_fit <- safe_read_csv(path(outputs, "bridge_diagnostics.csv"))
history  <- safe_read_wh("mart_nowcast_history")
disr     <- safe_read_wh("raw_disr_req_quarterly")

com_levels <- c("iron_ore", "coal")
```

## Headline: current-quarter nowcasts (Mt)

```{r headline, results='asis'}
if (nrow(nowcast) == 0) {
  cat("*No nowcast available this run.*")
} else {
  nc <- nowcast |>
    mutate(commodity = factor(commodity, levels = com_levels)) |>
    arrange(commodity)
  for (i in seq_len(nrow(nc))) {
    r <- nc[i, ]
    cat(sprintf(
      "- **%s** (%s): **%.1f Mt**  (80%% CI: %.1f – %.1f Mt).  Share observed: %.0f%%.\n",
      toupper(as.character(r$commodity)),
      format(r$quarter_end, "%b %Y"),
      r$point_estimate_Mt,
      r$lower_80, r$upper_80,
      100 * r$share_observed
    ))
  }
}
```

## Nowcast vs recent actuals

```{r actuals-plot, fig.width=9, fig.height=4.5}
if (nrow(disr) > 0 && nrow(nowcast) > 0) {
  hist <- disr |>
    mutate(quarter_end = as.Date(quarter_end),
           commodity   = factor(commodity, levels = com_levels)) |>
    filter(!is.na(commodity),
           quarter_end >= as.Date("2019-01-01")) |>
    arrange(commodity, quarter_end)

  nc <- nowcast |>
    mutate(quarter_end = as.Date(quarter_end),
           commodity   = factor(commodity, levels = com_levels))

  ggplot(hist, aes(quarter_end, tonnes_Mt)) +
    geom_line(colour = "grey40") +
    geom_point(colour = "grey40", size = 1) +
    geom_point(data = nc,
               aes(quarter_end, point_estimate_Mt),
               colour = "steelblue", size = 3) +
    geom_linerange(data = nc,
                   aes(x = quarter_end,
                       ymin = lower_95, ymax = upper_95,
                       y = point_estimate_Mt),
                   colour = "steelblue", alpha = 0.35, linewidth = 1.2) +
    geom_linerange(data = nc,
                   aes(x = quarter_end,
                       ymin = lower_80, ymax = upper_80,
                       y = point_estimate_Mt),
                   colour = "steelblue", alpha = 0.65, linewidth = 1.8) +
    facet_wrap(~ commodity, scales = "free_y") +
    labs(x = NULL, y = "Quarterly exports (Mt)",
         title = "Quarterly export volume: history (grey) + current nowcast (blue)") +
    theme_minimal()
}
```

## Bridge diagnostics

```{r diagnostics}
if (nrow(diag_fit) > 0) {
  diag_fit |>
    mutate(
      commodity     = factor(commodity, levels = com_levels),
      beta_tonnage  = sprintf("%.3f (se %.3f)", beta_tonnage, beta_tonnage_se),
      r_squared     = round(r_squared, 3),
      rmse_train    = round(rmse_train, 3),
      dw_stat       = round(dw_stat, 2),
      rmse_valid    = round(rmse_valid, 2),
      rmse_naive    = round(rmse_naive, 2),
      ratio_vs_naive = round(ratio_vs_naive, 2)
    ) |>
    arrange(commodity) |>
    select(commodity, n_obs, r_squared, rmse_train, dw_stat,
           beta_tonnage, rmse_valid, rmse_naive, ratio_vs_naive) |>
    knitr::kable(
      caption = "Bridge training fit (`n_obs` = training quarters) and walk-forward validation vs seasonal-random-walk. `ratio_vs_naive` < 1 means the model beats naive."
    )
}
```

## PortWatch tonnage trend

```{r tonnage-plot, fig.width=9, fig.height=3.5}
if (nrow(tonnage) > 0) {
  tonnage |>
    mutate(quarter_end = as.Date(quarter_end),
           commodity   = factor(commodity, levels = com_levels)) |>
    filter(!is.na(commodity),
           quarter_end >= as.Date("2019-01-01")) |>
    ggplot(aes(quarter_end, tonnage / 1e6, colour = commodity)) +
    geom_line() +
    geom_point(size = 1) +
    facet_wrap(~ commodity, scales = "free_y") +
    labs(x = NULL, y = "Quarterly PortWatch tonnage (Mt)",
         title = "PortWatch quarterly tonnage by commodity") +
    theme_minimal() +
    theme(legend.position = "none")
}
```

## Change since last run

```{r delta}
if (nrow(history) >= 2) {
  by_com <- history |>
    mutate(run_timestamp = as.POSIXct(run_timestamp)) |>
    arrange(commodity, run_timestamp) |>
    group_by(commodity) |>
    summarise(
      prior_point = dplyr::nth(point_estimate, -2L),
      now_point   = dplyr::last(point_estimate),
      delta       = now_point - prior_point,
      hours_since = as.numeric(difftime(
                      dplyr::last(run_timestamp),
                      dplyr::nth(run_timestamp, -2L),
                      units = "hours")),
      .groups = "drop"
    ) |>
    filter(!is.na(prior_point))
  if (nrow(by_com) > 0) {
    by_com |>
      mutate(prior_point = round(prior_point, 2),
             now_point   = round(now_point, 2),
             delta       = round(delta, 2),
             hours_since = round(hours_since, 1)) |>
      knitr::kable()
  } else {
    cat("*First run for every commodity -- no prior to compare against.*")
  }
} else {
  cat("*First run -- no prior nowcast in history.*")
}
```

## Methodology footnote

For each commodity $c \in \{\text{iron\_ore}, \text{coal}\}$, quarterly frequency:

$$\log V_{c,Q} = \beta_0 + \beta_1 \log T^{SA}_{c,Q} + \beta_2 \log V_{c,Q-1} + \varepsilon_{c,Q}$$

- $V_{c,Q}$ = DISR Resources & Energy Quarterly Table 16, physical tonnage (Mt) by commodity. Coal is metallurgical + thermal summed.
- $T^{SA}_{c,Q}$ = STL seasonally-adjusted quarterly PortWatch tonnage, summed across ports whose `commodity_class` matches $c$. Tanker tonnage at LNG-whitelisted ports routes to `lng` at ingestion but is then scoped out of the bridge panel (LNG was dropped 2026-04-21 after zero-correlation finding).
- Fit window: 2019-Q1 → 2023-Q4 (RHS-constrained by PortWatch coverage). Validation: 2024-Q1 onward.
- Newey-West HAC standard errors with `lag = 1` at quarterly frequency.
- Uncertainty bands: residual bootstrap, scaled by $\sqrt{1 - \text{share\_observed}}$ so bands collapse as the current quarter fills in.
