---
title: "Australian Resource Export Volumes — Quarterly Nowcast"
subtitle: "resourcetracker · per-commodity tonnage nowcast (iron_ore, coal)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    self_contained: true
    theme: cosmo
params:
  outputs_dir: "../../outputs"
  warehouse_dir: "../../data/warehouse"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(readr)
library(ggplot2)
library(fs)

outputs <- params$outputs_dir
wh_dir  <- params$warehouse_dir

safe_read_csv <- function(p) {
  if (!file_exists(p)) return(tibble())
  suppressMessages(read_csv(p, show_col_types = FALSE))
}
safe_read_wh <- function(name) {
  p <- path(wh_dir, paste0(name, ".rds"))
  if (!file_exists(p)) return(tibble())
  tibble::as_tibble(readRDS(p))
}

nowcast  <- safe_read_csv(path(outputs, "nowcast_current.csv"))
tonnage  <- safe_read_csv(path(outputs, "tonnage_quarterly.csv"))
diag_fit <- safe_read_csv(path(outputs, "bridge_diagnostics.csv"))
history  <- safe_read_wh("mart_nowcast_history")
disr     <- safe_read_wh("raw_disr_req_quarterly")

com_levels <- c("iron_ore", "coal")

# For each commodity in `nowcast`, find the most-recent prior row in
# `history` for the same quarter that isn't this run, and compute the
# point + 80% band delta. Returns a tibble with one row per commodity;
# delta cols are NA when no prior exists.
compute_wow_delta <- function(nowcast, history) {
  if (nrow(nowcast) == 0 || nrow(history) == 0) {
    return(tibble::tibble(commodity = character(),
                          delta_point = double(),
                          delta_pct = double(),
                          prior_run = as.POSIXct(NA),
                          hours_since = double()))
  }
  history <- history |>
    mutate(run_timestamp = as.POSIXct(run_timestamp),
           quarter_end   = as.Date(quarter_end))
  purrr::map_dfr(seq_len(nrow(nowcast)), function(i) {
    cur <- nowcast[i, ]
    cur_ts <- as.POSIXct(cur$run_timestamp)
    prior <- history |>
      filter(commodity == cur$commodity,
             quarter_end == as.Date(cur$quarter_end),
             run_timestamp < cur_ts) |>
      arrange(desc(run_timestamp)) |>
      slice(1)
    if (nrow(prior) == 0) {
      tibble::tibble(commodity   = cur$commodity,
                     delta_point = NA_real_,
                     delta_pct   = NA_real_,
                     prior_run   = as.POSIXct(NA),
                     hours_since = NA_real_)
    } else {
      d <- cur$point_estimate_Mt - prior$point_estimate
      tibble::tibble(
        commodity   = cur$commodity,
        delta_point = d,
        delta_pct   = 100 * d / prior$point_estimate,
        prior_run   = prior$run_timestamp,
        hours_since = as.numeric(difftime(cur_ts, prior$run_timestamp,
                                          units = "hours"))
      )
    }
  })
}

delta <- compute_wow_delta(nowcast, history)
```

## Top line

```{r top-line, results='asis'}
if (nrow(nowcast) == 0) {
  cat("*No nowcast available this run.*\n")
} else {
  nc <- nowcast |>
    mutate(commodity = factor(commodity, levels = com_levels)) |>
    arrange(commodity) |>
    left_join(delta, by = "commodity")
  for (i in seq_len(nrow(nc))) {
    r <- nc[i, ]
    change <- if (is.na(r$delta_point)) {
      "first nowcast for this quarter"
    } else {
      sign_lbl <- if (r$delta_point >= 0) "+" else ""
      sprintf("%s%.1f Mt (%s%.1f%%) vs the run %.0fh ago",
              sign_lbl, r$delta_point, sign_lbl, r$delta_pct, r$hours_since)
    }
    cat(sprintf(
      "- **%s** %s: **%.1f Mt** (80%% CI %.1f – %.1f). Quarter %.0f%% observed. Change: %s.\n",
      toupper(as.character(r$commodity)),
      format(as.Date(r$quarter_end), "%b %Y"),
      r$point_estimate_Mt,
      r$lower_80, r$upper_80,
      100 * r$share_observed,
      change
    ))
  }
}
```

## Headline: current-quarter nowcasts (Mt)

```{r headline, results='asis'}
if (nrow(nowcast) == 0) {
  cat("*No nowcast available this run.*")
} else {
  nc <- nowcast |>
    mutate(commodity = factor(commodity, levels = com_levels)) |>
    arrange(commodity)
  for (i in seq_len(nrow(nc))) {
    r <- nc[i, ]
    cat(sprintf(
      "- **%s** (%s): **%.1f Mt**  (80%% CI: %.1f – %.1f Mt).  Share observed: %.0f%%.\n",
      toupper(as.character(r$commodity)),
      format(r$quarter_end, "%b %Y"),
      r$point_estimate_Mt,
      r$lower_80, r$upper_80,
      100 * r$share_observed
    ))
  }
}
```

## Nowcast vs recent actuals

```{r actuals-plot, fig.width=9, fig.height=4.5}
if (nrow(disr) > 0 && nrow(nowcast) > 0) {
  hist <- disr |>
    mutate(quarter_end = as.Date(quarter_end),
           commodity   = factor(commodity, levels = com_levels)) |>
    filter(!is.na(commodity),
           quarter_end >= as.Date("2019-01-01")) |>
    arrange(commodity, quarter_end)

  nc <- nowcast |>
    mutate(quarter_end = as.Date(quarter_end),
           commodity   = factor(commodity, levels = com_levels))

  ggplot(hist, aes(quarter_end, tonnes_Mt)) +
    geom_line(colour = "grey40") +
    geom_point(colour = "grey40", size = 1) +
    geom_point(data = nc,
               aes(quarter_end, point_estimate_Mt),
               colour = "steelblue", size = 3) +
    geom_linerange(data = nc,
                   aes(x = quarter_end,
                       ymin = lower_95, ymax = upper_95,
                       y = point_estimate_Mt),
                   colour = "steelblue", alpha = 0.35, linewidth = 1.2) +
    geom_linerange(data = nc,
                   aes(x = quarter_end,
                       ymin = lower_80, ymax = upper_80,
                       y = point_estimate_Mt),
                   colour = "steelblue", alpha = 0.65, linewidth = 1.8) +
    facet_wrap(~ commodity, scales = "free_y") +
    labs(x = NULL, y = "Quarterly exports (Mt)",
         title = "Quarterly export volume: history (grey) + current nowcast (blue)") +
    theme_minimal()
}
```

## Bridge diagnostics

```{r diagnostics}
if (nrow(diag_fit) > 0) {
  diag_fit |>
    mutate(
      commodity     = factor(commodity, levels = com_levels),
      beta_tonnage  = sprintf("%.3f (se %.3f)", beta_tonnage, beta_tonnage_se),
      r_squared     = round(r_squared, 3),
      rmse_train    = round(rmse_train, 3),
      dw_stat       = round(dw_stat, 2),
      rmse_valid    = round(rmse_valid, 2),
      rmse_naive    = round(rmse_naive, 2),
      ratio_vs_naive = round(ratio_vs_naive, 2)
    ) |>
    arrange(commodity) |>
    select(commodity, n_obs, r_squared, rmse_train, dw_stat,
           beta_tonnage, rmse_valid, rmse_naive, ratio_vs_naive) |>
    knitr::kable(
      caption = "Bridge training fit (`n_obs` = training quarters) and walk-forward validation vs seasonal-random-walk. `ratio_vs_naive` < 1 means the model beats naive."
    )
}
```

## PortWatch tonnage trend

```{r tonnage-plot, fig.width=9, fig.height=3.5}
if (nrow(tonnage) > 0) {
  base <- tonnage |>
    mutate(quarter_end = as.Date(quarter_end),
           commodity   = factor(commodity, levels = com_levels)) |>
    filter(!is.na(commodity),
           quarter_end >= as.Date("2019-01-01"))

  # PortWatch coverage stops at max(obs_date), so the most-recent quarter
  # has a raw `tonnage` that under-represents the full quarter. Split:
  # plot completed quarters as the solid history line; show the partial
  # quarter as a hollow point at its extrapolated full-quarter value,
  # connected by a dashed segment.
  complete <- filter(base, is_complete)
  partial  <- filter(base, !is_complete, !is.na(tonnage_extrapolated))

  # Bridge segment from last-complete point to partial-extrapolated point
  bridge_seg <- if (nrow(complete) > 0 && nrow(partial) > 0) {
    complete |>
      group_by(commodity) |>
      filter(quarter_end == max(quarter_end)) |>
      ungroup() |>
      transmute(commodity,
                x = quarter_end, y = tonnage / 1e6) |>
      left_join(
        partial |> transmute(commodity,
                             xend = quarter_end,
                             yend = tonnage_extrapolated / 1e6),
        by = "commodity"
      ) |>
      filter(!is.na(.data$yend))
  } else {
    NULL
  }

  p <- ggplot() +
    geom_line(data = complete,
              aes(quarter_end, tonnage / 1e6, colour = commodity)) +
    geom_point(data = complete,
               aes(quarter_end, tonnage / 1e6, colour = commodity),
               size = 1)
  if (!is.null(bridge_seg) && nrow(bridge_seg) > 0) {
    p <- p +
      geom_segment(data = bridge_seg,
                   aes(x = x, y = y, xend = xend, yend = yend,
                       colour = commodity),
                   linetype = "dashed")
  }
  if (nrow(partial) > 0) {
    p <- p +
      geom_point(data = partial,
                 aes(quarter_end, tonnage_extrapolated / 1e6,
                     colour = commodity),
                 size = 3, shape = 21, fill = "white", stroke = 1.2)
  }
  p +
    facet_wrap(~ commodity, scales = "free_y") +
    labs(x = NULL, y = "Quarterly PortWatch tonnage (Mt)",
         title = "PortWatch quarterly tonnage by commodity",
         caption = paste(
           "Solid: completed quarters (raw observed PortWatch sum).",
           "Hollow point: current partial quarter extrapolated to a",
           "full-quarter total using the same logic the bridge consumes",
           "(observed past months + day-of-month scale-up for the",
           "current month + seasonal pace for unobserved future months)."
         )) +
    theme_minimal() +
    theme(legend.position = "none",
          plot.caption   = element_text(size = 7, hjust = 0))
}
```

## Change since last run

```{r delta, results='asis'}
# Reuse the `delta` tibble from the Top line; join back to nowcast to
# pull the now-point so this section stands alone.
if (nrow(nowcast) == 0 || nrow(delta) == 0) {
  cat("*No nowcast available this run.*\n")
} else if (all(is.na(delta$delta_point))) {
  cat("*First run for every commodity (this quarter) -- no prior to compare against.*\n")
} else {
  nowcast |>
    select(commodity, quarter_end, now_Mt = point_estimate_Mt) |>
    left_join(delta, by = "commodity") |>
    filter(!is.na(delta_point)) |>
    mutate(
      commodity   = factor(commodity, levels = com_levels),
      prior_Mt    = round(now_Mt - delta_point, 2),
      now_Mt      = round(now_Mt, 2),
      delta_Mt    = round(delta_point, 2),
      delta_pct   = round(delta_pct, 2),
      hours_since = round(hours_since, 1)
    ) |>
    arrange(commodity) |>
    select(commodity, quarter_end, prior_Mt, now_Mt, delta_Mt, delta_pct,
           hours_since) |>
    knitr::kable(caption = "Same-quarter run-over-run delta. Drops rows for any commodity whose current quarter has no prior nowcast.")
}
```

## Methodology footnote

For each commodity $c \in \{\text{iron\_ore}, \text{coal}\}$, quarterly frequency:

$$\log V_{c,Q} = \beta_0 + \beta_1 \log T^{SA}_{c,Q} + \beta_2 \log V_{c,Q-1} + \varepsilon_{c,Q}$$

- $V_{c,Q}$ = DISR Resources & Energy Quarterly Table 16, physical tonnage (Mt) by commodity. Coal is metallurgical + thermal summed.
- $T^{SA}_{c,Q}$ = STL seasonally-adjusted quarterly PortWatch tonnage, summed across ports whose `commodity_class` matches $c$. Tanker tonnage at LNG-whitelisted ports routes to `lng` at ingestion but is then scoped out of the bridge panel (LNG was dropped 2026-04-21 after zero-correlation finding).
- Fit window: 2019-Q1 → 2023-Q4 (RHS-constrained by PortWatch coverage). Validation: 2024-Q1 onward.
- Newey-West HAC standard errors with `lag = 1` at quarterly frequency.
- Uncertainty bands: residual bootstrap, scaled by $\sqrt{1 - \text{share\_observed}}$ so bands collapse as the current quarter fills in.
